---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = TRUE,
  comment = "#>"
)
```

The purpose of `rmaven` is to simplify the process of including Java libraries
into R code for use by `rJava`. Currently the `rJava` infrastructure assumes jar
files of compiled code, and all the dependencies thereof, are available locally
and known to the programmer. This works well if the `rJava` code is part of a
package, but less well if the user is programming in a `REPL` loop, and simply
wants to use a Java library.

Even when packaging Java libraries with `rJava` code into an R package, the
transitive dependencies of a moderately complicated bit of Java code can be
complex, and distributing the `jar` files associated with that burdensome,
particularly when the underlying libraries are updated. The bundling of jar
files also may exceed the stringent CRAN policies for allowable package sizes,
requiring a complex distribution mechanism for the dependencies.

To solve this `rmaven` manages some or all of this problem for you, by
integrating Java's standard build and dependency management tool Apache Maven
into R. This allows Java libraries used within R to be specified as `Maven`
artifacts and transitive dependencies downloaded dynamically from Maven
repositories. `rmaven` is pure R with minimal dependencies of its own, apart
from the existence of a Java runtime (`JRE` or `JDK`) installed on the system,
and the location known to R. It doesn't even require `rJava` (although without
it there doesn't seem much point).

```{r setup}
library(rmaven)
start_jvm()

# The following 3 lines clears the local rmaven cache and is here to 
# create a clean slate before executing the rest of the vignette.
# In normal use you would not call this in a non-interactive setting
opts = options("rmaven.allow.cache.delete"=TRUE)
clear_rmaven_cache()
options(opts)

```

# Download jar files using Maven

When using `rmaven` the first step to using a Java library in R is to fetch it
from the Maven repositories. The artifact coordinates (i.e. `groupId`,
`artifactId` and `version`) of an specific Java library are easily found on the
internet. Maven manages bootstrapping the download of libraries.

```{r}
# cache apache commons lang3 
dynamic_classpath = resolve_dependencies(
  groupId = "org.apache.commons", 
  artifactId = "commons-lang3", 
  version="3.12.0"
)

sprintf("Locally cached classpath of commons-lang3: %s",dynamic_classpath)

# to get the one jar file many others are donloaded and cached as part of
# bootstrapping the Java build tools
sprintf("Jar files found in repository: %1.0f", 
        fs::dir_ls(get_repository_location(),recurse = TRUE,glob="*.jar") %>% length())
```

In Java development Maven manages a local repository in the `.m2/repository/`
subdirectory of the users home. In `rmaven` however changing directories in the
user space conflicts with CRAN policies. As the purpose of this library is to
support Java use in R, it makes sense to move the `.m2/repository/` cache
directory to a CRAN approved location, however this can be changed through
configuration (see `set_repository_location(...)` function).

# Caching of Java libraries

As we saw above `rmaven` bootstraps loading of Maven itself and any plugins
required. When doing this the code above triggers a significant download. Maven
is good at handling the caching of downloads, and sharing reused artifacts.
Repeated invocation of `rmaven` will not require additional downloads.

```{r}
# does not need additional downloads
resolve_dependencies(
  groupId = "org.apache.commons", 
  artifactId = "commons-lang3", 
  version="3.12.0"
)

sprintf("Unchanged locally cached classpath of commons-lang3: %s",dynamic_classpath)

# no additional downloads once cache populated
sprintf("Jar files found in repository: %1.0f", 
        fs::dir_ls(get_repository_location(),recurse = TRUE,glob="*.jar") %>% length())
```

# Using local repository Jar files in `rJava`

The second step is to add jars from the local `.m2` repository to the `rJava`
class path, and using `rJava`, create an R interface to the Java class you want
to use (in this case the `StringUtils` class).

```{r}
rJava::.jaddClassPath(dynamic_classpath)
StringUtils = rJava::J("org.apache.commons.lang3.StringUtils")
```

Finally you can call the static Java method, using `rJava`, in this case
`StringUtils.rotate(String s, int distance)`.

```{r}
StringUtils$rotate("ABCDEF",3L)
```

# Java compilation

Apart from using Maven to download manage dependencies, it also allows us to
compile Java code and integrate dependencies, specified in a `pom.xml` file. By
including Java code in an R project and using Maven to compile code we raise the
possibility of integrating Java and R code in a manner similar to the way `RCpp`
and `C++`.

## Step 0: write some java code 

As part of your package write some Java code including a `pom.xml` Maven build
file, such as the `TestCass`, in the example code bundled with this package.

```{r}
# find the java source code example bundled with this package
source_directory = system.file("testdata/test-project",package = "rmaven")
fs::dir_tree(source_directory, invert=TRUE, glob = "*/target/*")
```

## Step 1: compile the jar

We use `rmaven` to compile the java source code. This could possibly be done in
an `.onLoad()` package function, if you were developing a package which includes
Java source code:

```{r}
# In this configuration rmaven will compile a single `fat-jar`
compiled_jar = compile_jar(source_directory, with_dependencies = TRUE)

sprintf("Location of compiled jar from test source code in this repository: %s",compiled_jar)

# More plugins are required to perform the compilation
sprintf("Jar files found in repository: %1.0f", 
        fs::dir_ls(get_repository_location(),recurse = TRUE,glob="*.jar") %>% length())

```

## Step 2: use the jar

Once compiled we can use the Java code within R in a very similar manner as before

```{r}
rJava::.jaddClassPath(compiled_jar)
TestClass = rJava::J("org.example.TestClass")
TestClass$sayHelloWorld()
```

# Options and configuration

Various options exist to control behaviour of the library. N.B. None of the
options in this section are executed as part of the vignette, they are just here
to document the options.

Set the level of information returned by Maven. Maven can be very verbose so 
defaulting `rmaven` to "quiet" mode is often a good idea.

```R
options("rmaven.quiet"=TRUE)
```

Set the default repository URLs for fetching artifacts.
these can also be specified as options for the `fetch_artifact()` and
`copy_artifact()` functions.
E.g. to restrict to Maven central:

```R
options("rmaven.default_repos" = c("https://repo1.maven.org/maven2/"))
# N.b. plan is to integrate this with settings.xml
```

Manually set `JAVA_HOME` for Maven to use. The home directory of the Java
installation is detected automatically by firstly checking this option, then the
`JAVA_HOME` environment variable, then the `LD_LIBRARY_PATH` environment
variable, then by asking `rJava`.

```R
options("rmaven.java_home"="/usr/lib/jvm/java-9-oracle")
```

Provide custom options to the JVM on startup, e.g. enabling profiling, 
this needs to be set before `start_jvm()` to have any effect: 

```R
options("java.parameters"=c("-Xprof","-Xrunhprof"))
```

Enable debug mode option has the dual effect of:
starting the JVM in debug mode (JVM flags: 
`-Xdebug -Xrunjdwp:transport=dt_socket,address=8998,server=y,suspend=n` 
), and asking Maven to print debugging output (Maven flags: `-X -e`). 
This needs to be set before `start_jvm()` to have any effect.

```R
options("rmaven.debug"=TRUE)
```

Configure location of the local m2 repository, e.g. to use the Java default 
package location:

You might want to do this if you are a Java developer and 
you have a maven cache already populated with locally built Java libraries
from a Java IDE and you are working to develop an R package pending deployment.
If you using `rmaven` for dependency management in a package and
you intend to submit it to CRAN, altering this setting
for the `rmaven` repository location will cause issues during CRAN submission
as the resulting package will touch the userspace and hence violate CRAN's 
policies.

```R
# THIS IS NOT RUN (and nore are any of the other options documented 
# in this section)
# options("rmaven.m2.repository"="~/.m2/repository/")
```

Allow `rmaven` to programmatically delete all cached files. 
This will involve a lot of downloading from maven central and generally only 
required on a development machine where you want to make sure everything runs 
from scratch (like in this vignette).

```R
options("rmaven.allow.cache.delete"=TRUE)
# combined with clear_rmaven_cache()
```

# What else could you do?

* Run ant tasks. - using the `antrun` plugin
* Initialize production web servers and 'APIs'. - using Jetty or Spring.
* run any other maven plugin goal, for example:

```{r}
# Here we execute a maven help goal
execute_maven("help:system")

```


# Further work

* Maven shade plugin
* Simplify multiple dependencies 
* Support creation of `pom.xml` files from R
* Tools to maintain maven `settings.xml`
