---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = TRUE,
  comment = "#>"
)
```

The purpose of `rmaven` is to simplify the process of including Java libraries into R code for use by `rJava`. Currently the 
`rJava` infrastructure assumes jar files of compiled code, and all the dependencies thereof, are available locally and known to the programmer. This 
works well if the `rJava` code is part of a package, but less well if the user is programming in a `REPL` loop, and simply wants to use a Java library. 

Even when packaging Java libraries with `rJava` code into an R package, the transitive dependencies of a moderately complicated bit of Java code can 
be complex, and distributing the `jar` files associated with that burdensome, particularly when the underlying libraries are updated. The bundling of jar files 
also may exceed the stringent CRAN policies for allowable package sizes, requiring a complex distribution mechanism for the dependencies.

To solve this `rmaven` manages some or all of this problem for you, by integrating Java's standard build and dependency management tool Apache Maven into R. 
This allows Java libraries used within R to be specified as `Maven` artifacts and transitive dependencies downloaded dynamically from Maven repositories. 
`rmaven` is pure R with minimal dependencies of its own, apart from the 
existence of a Java runtime (`JRE` or `JDK`) installed on the system, and the location known to R. It doesn't even require `rJava` (although without it there
doesn't seem much point).

```{r setup}
library(rmaven)
start_jvm()

opts = options("rmaven.allow.cache.delete"=TRUE)
clear_rmaven_cache()
options(opts)
```

# Download jar files using Maven

When using `rmaven` the first step to using a Java library in R is to fetch it from the Maven repositories. The artifact coordinates (i.e. `groupId`, `artifactId` and `version`) of
an specific Java library are easily found on the internet. Maven manages bootstrapping the download of libraries. 

```{r}
dynamic_classpath = resolve_dependencies(
  groupId = "org.apache.commons", 
  artifactId = "commons-lang3", 
  version="3.12.0",
  nocache = TRUE
)
dynamic_classpath
```


In Java development Maven manages a local repository in the `.m2/repository/` 
subdirectory of the users home. In `rmaven` however changing directories in the user space conflicts with CRAN policies. As the purpose of this library is to support Java use in R, it makes sense
to move the `.m2/repository/` cache directory to a CRAN approved location, however this can be changed through configuration (see `set_repository_location(...)` function).

# Caching of Java libraries

`rmaven` bootstraps loading of Maven itself and any plugins required. When doing this the code above triggers a significant download. Maven is good at handling the caching of downloads, and sharing reused artifacts. Repeated invocation of `rmaven` will not require additional downloads.

```{r}
# does not need additional downloads
resolve_dependencies(
  groupId = "org.apache.commons", 
  artifactId = "commons-lang3", 
  version="3.12.0"
)
```

# Using local repository Jar files in `rJava`

The second step is to add jars from the local `.m2` repository to the `rJava` class path, and using `rJava`, create an R interface to the Java class you want to use (in this case the `StringUtils` class). 

```{r}
rJava::.jaddClassPath(dynamic_classpath)
StringUtils = rJava::J("org.apache.commons.lang3.StringUtils")
```

Finally you can call the static Java method, using `rJava`, in this case `StringUtils.rotate(String s, int distance)`.

```{r}
StringUtils$rotate("ABCDEF",3L)
```

# Java compilation

Apart from using Maven to download manage dependencies, it also allows us to compile Java code and integrate dependencies, specified in a `pom.xml` file. By including Java code in an R project and using Maven to compile code we raise the possibility of integrating Java and R code in a manner similar to the way `RCpp` and `C++`.

## Step 0: write some java code 

As part of your package write some Java code including a `pom.xml` Maven build file, such as the `TestCass`, in the example code bundled with this package.

```{r}
# find the example bundled with this package
source_directory = system.file("testdata/test-project",package = "rmaven")
fs::dir_tree(source_directory, invert=TRUE, glob = "*/target/*")
```

## Step 1: compile the jar

We use `rmaven` to compile the java source code. This could possibly be done in an `.onLoad()` package function:

```{r}
# In this configuration rmaven will compile a single `fat-jar`
compiled_jar = rmaven::compile_jar(source_directory, with_dependencies = TRUE)
compiled_jar
```

## Step 2: use the jar

Once compiled we can use the Java code within R in a very similar manner as before

```{r}
rJava::.jaddClassPath(compiled_jar)
TestClass = rJava::J("org.example.TestClass")
TestClass$sayHelloWorld()
```

# What else could you do?

* Run ant tasks. - using the `antrun` plugin
* Initialize production web servers and 'APIs'. - using Jetty or Spring.
* run any other maven plugin goal:

```{r}
execute_maven("help:system")
```


# Further work

* Maven shade plugin
* Simplify multiple dependencies 
* Support creation of `pom.xml` files from R
* Tools to maintain maven `settings.xml`
